---
title: "Who Gets What: How Population and Income Shape Toronto’s Budget Decisions"
subtitle: "Understanding the Impact of Demographics and Economic Factors on City Spending Across 25 Wards"
author: 
  - Aviral Bhardwaj
thanks: "Code and data are available at: https://github.com/Aviral-03/Toronto-WardWide-Budget-Analysis"
date: today
date-format: long
abstract: "This study investigates how population density and average household income affect budget allocations across Toronto’s 25 wards. Using data from the city's 2023-2032 Capital Budget Plan and 2021 Ward Profiles, we analyze spending patterns in key areas like health and safety, infrastructure, and community services. Our findings reveal that higher population densities do not guarantee proportionally greater investments in essential services, while wealthier wards often receive larger capital expenditures. These insights highlight potential disparities in resource distribution, emphasizing the need for more equitable urban budget planning to address diverse community needs."
format: pdf
number-sections: true
toc: true
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false

library(tidyverse)
library(opendatatoronto)
library(readxl)
library(styler)
library(dplyr)
library(arrow)
library(ggrepel)
library(knitr)
library(kableExtra)
library(patchwork)
library(sf)
library(lintr)
```

```{r}
#| include: false
#| warning: false
#| message: false

ward_data <- read.csv("../data/analysis_data/cleaned_ward_data.csv")
capital_data <- read.csv("../data/analysis_data/cleaned_capital_data.csv")

```

# Introduction {#sec-intro}

In urban governance, equitable allocation of municipal budgets is crucial for addressing city's diverse social, economic, and infrastructural needs. This challenge is particularly acute in Toronto, one of Canada's largest and most diverse cities, where 25 wards compete for limited resources with ever growing population. Effective allocation requires a careful balance, but understanding how demographic and economic factors influence these decisions remains an under-explored area. This study addresses this gap by investigating the relationship between these factors and Toronto's budget allocations, with implications for more informed, data-driven urban planning. Specifically, we aim to answer the question: To what extent do demographic and economic factors contribute to an equitable distribution of municipal resources across Toronto’s wards?

Historical budget data reveals a pattern of uneven investment, particularly in areas with high population density and lower-income residents, which has contributed to growing disparities in service delivery and infrastructure quality [@ferguson2024compromised]. Following her election in 2023, Mayor Olivia Chow prioritized resolving Toronto’s housing crisis, improving transportation, and addressing social inequities [@toronto_2024_budget]. The city’s 2024 budget reflects these priorities, with notable investments in affordable housing, transit, and community safety initiatives. Simultaneously, the Chief Financial Officer underscored the importance of long-term infrastructure development, highlighting the city’s 10-year capital plan [@toronto_2024_budget]. However, it remains unclear to what extent these priorities translate into equitable budget allocations across Toronto's wards, particularly in health and safety or growth-related expenditures.

To explore this issue, this study leverages multiple datasets from the City of Toronto’s Open Data Portal, including demographic information from the 2021 Ward Profiles and budget data from the 2023-2032 Capital Budget Plan. Our analysis focuses on two key demographic factors—population density and average household income—and examines their causal effect on budget allocations across categories such as Health and Safety, Growth-Related Expenditures, and Service Improvement. By focusing on these variables, we aim to uncover whether Toronto’s budget distribution aligns with the priorities of Mayor Chow and the city’s Chief Financial Officer.

Our findings reveal that wards with higher population densities do not consistently receive a proportionate share of investments in essential areas like health and safety. While wealthier wards tend to receive more substantial capital expenditures, however, the budget does not show a strong bias in their favor overall for other categories. These results raise concerns about potential disparities in resource distribution and highlight the need for more equitable budget planning. This paper is structured as follows: [@sec-data] discusses the data and methodologies used, followed by discussions of the relationship between variables of interest, and [@sec-discussion] explores the broader implications for urban governance and fiscal policy.

# Data {#sec-data}

The raw data was sourced from the City of Toronto's Open Data Portal using the `opendatatoronto` [@openDataToronto] package. Three data sets were downloaded: `2023 Ward Profiles (25-Ward Model)`[@toronto_ward_profiles], `2023-2032 Capital Budget and Plan Details`[@toronto_budget_plan], and `City Wards`[@toronto_city_wards]. The data, provided in Excel and CSV formats, was cleaned and analyzed using R [@citeR] programming language. The `readxl` [@readxl] package was used for reading Excel files. Other R packages used include `tidyverse` [@tidyverse], `styler` [@styler], and `dplyr` [@dplyr] for creating tables. The `ggplot2` [@ggplot2] and `kableExtra` [@kableExtra] were used for data visualization and table formatting. The `patchwork` [@patchwork] package was used for combining multiple plots, and `sf` [@sf] for spatial data analysis.

## Measurement
Our research question and estimand analyse the relationship between key demographic and economic factors—specifically, population and average household income—and the budget allocation across various categories in Toronto's 25 wards. Population is a critical demographic indicator, representing the number of residents in each ward, while average household income reflects economic well-being, influencing access to resources and quality of life [@Schaeffer]. Therefore, these two variables of interest were choosen for the analysis. Our primary aim is to estimate the impact of these variables on budget allocations for different categories, including Health and Safety, Growth-Related expenditures, State of Good Repair, and Service Improvement and Enhancement.

Population density is expected to affect the demand for services and infrastructure, while average household income may shape how resources are distributed across wards, reflecting broader economic disparities. The estimand is the causal effect of these factors on the allocation of funds, and the estimator will quantify how shifts in population and income levels influence budgetary decisions. By understanding this relationship, we aim to offer insights into how demographic and economic variables drive municipal spending patterns in different categories across the city.

## Ward Profiles (25-Ward Model) {#sec-ward-profiles}

The 2021 Ward Profiles [@toronto_ward_profiles], based on the 25-Ward model were provided by City Planning. These profiles included census data from the 2021, 2016, and 2011 Census of Population, covering demographic, social, and economic information for each ward in Toronto. These variables were collected through methods including online responses, mailed questionnaires, the Census Help Line, and enumerators [@nhs_canada]. 

This data-set was included in this analysis to provide insights into the population and average household income for each ward, providing insights into the city's socioeconomic landscape. 25-Ward model was used instead of the 44-Ward model as it was the most recent data available at the time of analysis and matched the Capital Budget data.

The data was stored in an Excel workbook with multiple tabs, but for this analysis, we used the first tab, `2016 Census One variable`, which contains data for all 25 wards (Ward 1, Ward 2, ..., Ward 25). After cleaning, the data was saved in CSV and Parquet formats, with the following columns:

-   `ward_id`: unique identifier for each ward,
-   `ward`: ward name,
-   `population`: total population,
-   `income`: average household income.

The ward names were manually entered into the cleaned data and the `ward_id` was set as the row name. A sample of the data can be seen in [@tbl-ward-profile-data].

```{r}
#| echo: false
#| message: false
#| label: tbl-ward-profile-data
#| tbl-cap: Sample of Cleaned Toronto Ward Profile Data
#| tbl-pos: "h"

# Remove the first column
ward_data |> 
  select(-1) |> 
  head(5) |>
  kable(col.names = c("Ward ID", "Ward Name", "Population", 
                      "Average Household Income"), 
        booktabs = TRUE, 
        align = "c")
```
## Capital Budget and Plan Details {#sec-capital-budget}

The 2023-2032 Capital Budget and Plan Details dataset [@toronto_budget_plan], published by Toronto's Financial Planning Division and updated on May 1, 2024, provides the 10-year capital budget breakdown for the city’s wards. This budget is developed through a collaborative process, where city staff prepare an initial draft, which is then reviewed by the Budget Committee. Input is solicited from Toronto residents and businesses, and subsequently, the Mayor presents the finalized budget proposal by February 1. City Council reviews and considers this budget within 30 days [@toronto_2024_budget]. Through this process, City Council makes decisions about City services, programs and infrastructure that impact the Toronto community.

This data set covers allocations for fixed assets such as land, buildings, equipment, and new facility construction. We used it to analyze the distribution of capital budget allocations across various categories and wards. The 2023-2032 time frame was selected to align with Ward Profiles data and to reflect the planning relevance following the 2022 Municipal Elections.

Key columns include `Project Name`, yearly budget allocations from 2023 to 2032, `Ward Number`, `Ward`, `Category`, and `Total 10 Year` (Sum of Year 1 to 10), where the budget is in thousands of dollars (e.g., 10 = \$10,000). However,  [@tbl-capital-budget-data] shows a sample of the cleaned data and our variables of interest:

-   `Total 10-year capital budget` allocated to each ward
-   `Category of the Capital Budget`, including fields like `Service Improvement and Enhancement`, `Growth Related`, and `State of Good Repair`.

Rows with `CW` (city-wide budget) were removed since they were applicable to all wards and to avoid large values, keeping only rows specific to each ward. The cleaned data was saved in CSV and Parquet formats for further analysis.

```{r}
#| echo: false
#| message: false
#| label: tbl-capital-budget-data
#| tbl-cap: Sample of Cleaned Toronto Capital Budget Data
#| tbl-pos: "h"

capital_data |>
  select(-1) |>
  head(5) |>
  kable(col.names = c("Ward ID", "Ward Name", "Category", 
                  "Total 10-Year Budget (in 000s)"),
    booktabs = TRUE,
    align = "c"
  )
```

## City Wards {#sec-ward-names}

The City Wards dataset [@toronto_city_wards], published by the City Clerk's Office and last updated on July 22, 2024, contains geographical information about each ward, including the ward ID, ward name, and ward boundary. These ward boundaries were decided as a part of ` Bill 5, Better Local Government Act` in 2018, reducing the number of wards from 47 to 25 [@toronto_city_wards].

This dataset, effective January 1, 2024, was used to map the `ward_id` to the ward name in the cleaned data. Key columns include:

-   `ward_id`: unique identifier for each ward,
-   `ward`: ward name,
-   `ward_boundary`: geographical boundary of the ward.

The ward names were mapped to the `ward_id` and integrated with the Ward Profiles [@sec-ward-profiles] and Capital Budget data sets [@sec-capital-budget] to create the final data set for analysis. This dataset was not used directly in the analysis but was essential for mapping the ward names to the ward IDs in the cleaned data.

```{r}
#| include: false
#| warning: false
#| message: false

combined_data <- merge(ward_data, capital_data, by = "ward_id") |>
  select("ward_id", "ward.x", "population", "average_household_income", 
         "category", "total_10_year") |>
  rename("ward" = "ward.x")
```

```{r}
#| echo: false
#| message: false
#| warning: false
#| label: fig-toronto-map-population
#| fig-cap: "Map of Toronto highlighting the population and income densities by ward"
#| fig.pos: 'H'
#| fig.height: 5

# URL to the zip file
url <- "https://ckan0.cf.opendata.inter.prod-toronto.ca/dataset/5e7a8234-f805-43ac-820f-03d7c360b588/resource/35f67d86-cfc8-4483-8d77-50d035b010d9/download/25-ward-model-december-2018-wgs84-latitude-longitude.zip"

# Temporary file to store the downloaded zip
temp_zip <- tempfile(fileext = ".zip")

# Download the zip file
download.file(url, temp_zip, mode = "wb")

# Unzip the file to a temporary directory
temp_dir <- tempdir()
unzip(temp_zip, exdir = temp_dir)

shapefiles <- list.files(temp_dir, pattern = "\\.shp$", full.names = TRUE)

toronto_map <- st_read(shapefiles[1], quiet = TRUE)

# Clean up the downloaded zip file
unlink(temp_zip)

# Merge analysis_data with toronto_map by ward_id
toronto_map <- merge(toronto_map, ward_data, 
                     by.x = "AREA_NAME", by.y = "ward") |>
  rename("income" = "average_household_income")

color_scale <- scale_fill_gradient(low = "lightblue", high = "darkblue")


# Plot with facet_wrap for average household income and population
plot1 <-ggplot() +
  geom_sf(data = toronto_map, aes(fill = population)) +
  geom_sf_text(data = toronto_map, aes(label = ward_id), 
               size = 2, color = "white") +
  color_scale +
  theme_minimal() +
  labs(x = "Longitude", y = "Latitude") +
  theme(legend.position = "right")


plot2 <- ggplot() +
  geom_sf(data = toronto_map, aes(fill = income)) +
  geom_sf_text(data = toronto_map, aes(label = ward_id), size = 2, color = "white") +
  color_scale +
  theme_minimal() +
  labs(x = "Longitude", y = "Latitude") +
  theme(legend.position = "right")

plot1 / plot2


```

## Ward Profiles Results {#sec-ward-profile-results}

The City of Toronto is divided into 25 wards, and the 2021 census data highlights significant disparities in population and average household income among them. As shown in [@fig-toronto-map-population], Spadina-Fort York (Ward 10) and Etobicoke-Lakeshore (Ward 3) have the highest population densities, with 135,400 and 139,920 residents, respectively. In contrast, Don Valley West (Ward 15) has a lower population density of 101,025 but boasts the highest average household income at \$224,800.

Population distribution reveals a concentration in the western part of the city and downtown areas. Notably, wards like Etobicoke Centre (Ward 2) and Etobicoke-Lakeshore (Ward 3), situated farther from downtown, have lower costs of living and correspondingly lower average household incomes, such as Etobicoke North (Ward 2) with \$95,200.

An interesting pattern emerges around Don Valley West (Ward 15), which has the highest average household income. Wards such as University-Rosedale (Ward 11), Toronto-St. Paul's (Ward 12), and Eglinton-Lawrence (Ward 8) are clustered together, indicating a geographical correlation among high-income areas. Conversely, the wards with the highest population densities do not overlap with those that have the highest average household incomes, highlighting distinct socioeconomic patterns within the city.

This analysis will further explore the relationships between average household income, population, and budget allocations across various categories, with a focus on wards exhibiting higher average household incomes and population densities.

## Relationship between Average Household Income & Population with Budget Allocation for Health and Safety

```{r}
#| echo: false
#| message: false
#| warning: false
#| fig.cap: "Health and Safety Budget by Ward"
#| label: fig-health-safety-budget
#| fig-pos: "h"
#| fig.width: 10
#| fig.height: 5

# Plot for Average Household Income by Ward
combined_data |>
  filter(category == "Health and Safety") |>
  select(ward_id, average_household_income, population, total_10_year) |>
  pivot_longer(cols = c("average_household_income", "population"),
               names_to = "variable",
               values_to = "value") |>
  ggplot(aes(x = value, y = total_10_year, label = ward_id)) +
  geom_point() +
  geom_text_repel(hjust = -0.5, vjust = 0.5) +
  geom_smooth(method = "lm", color = "blue", se = TRUE) +
  facet_wrap(~variable, scales = "free", labeller = as_labeller(c(
    average_household_income = "Average Household Income",
    population = "Population"
  ))) +
  theme_minimal() +
  labs(y = "Total 10-Year Budget (in 000s)")
```

[@fig-health-safety-budget] illustrates the relationship between average household income, population, and the total 10-year budget allocation for Health and Safety projects. The linear regression lines indicate a positive correlation, suggesting that higher average household incomes and larger populations are associated with increased budget allocations for Health and Safety.

As noted in [@sec-ward-profile-results], Ward 15 (Don Valley West) has the highest average household income, and surrounding wards also exhibit elevated incomes. This trend is reflected in the budget allocations, with Ward 11 receiving a higher budget for Health and Safety projects compared to other wards. This implies that areas with greater average household incomes tend to receive more funding for such initiatives.

Ward 10 (Spadina-Fort York), despite its high population density, ranks second in budget allocation. This indicates a prioritization of higher-income areas for funding, even though Ward 10 has a significant population. The lower budget allocation for Ward 10 compared to Ward 11 suggests a strategic focus on income levels rather than solely on population density.

It is important to note that some data points are missing in the plot for other wards, which may result from incomplete data or other factors. Further analysis is warranted to better understand the intricate relationships among average household income, population, and budget allocations for Health and Safety projects.

## Relationship between Average Household Income & Population with Budget Allocation for Growth Related Projects

```{r}
#| echo: false
#| message: false
#| warning: false
#| label: fig-growth-related-budget
#| fig.cap: "Growth Related Budget by Ward"
#| fig-pos: "h"
#| fig.width: 10
#| fig.height: 5

# Plot for Average Household Income by Ward
combined_data |>
  filter(category == "Growth Related") |>
  select(ward_id, average_household_income, population, total_10_year) |>
  pivot_longer(cols = c("average_household_income", "population"),
               names_to = "variable",
               values_to = "value") |>
  ggplot(aes(x = value, y = total_10_year, label = ward_id)) +
  geom_point() +
  geom_text_repel(hjust = -0.5, vjust = 0.5) +
  geom_smooth(method = "lm", color = "blue", se = TRUE) +
  facet_wrap(~variable, scales = "free", labeller = as_labeller(c(
    average_household_income = "Average Household Income",
    population = "Population"
  ))) +
  theme_minimal() +
  labs(y = "Total 10-Year Budget (in 000s)")

```

```{r}
#| echo: false
#| message: false
#| warning: false
#| fig.cap: "State of Good Repair Budget by Ward"
#| label: fig-state-good-repair-budget
#| fig-pos: "h"
#| fig.width: 10
#| fig.height: 5

# Plot for Average Household Income by Ward
combined_data |>
  filter(category == "State of Good Repair") |>
  select(ward_id, average_household_income, population, total_10_year) |>
  pivot_longer(cols = c("average_household_income", "population"),
               names_to = "variable",
               values_to = "value") |>
  ggplot(aes(x = value, y = total_10_year, label = ward_id)) +
  geom_point() +
  geom_text_repel(hjust = -0.5, vjust = 0.5) +
  geom_smooth(method = "lm", color = "blue", se = TRUE) +
  facet_wrap(~variable, scales = "free", labeller = as_labeller(c(
    average_household_income = "Average Household Income",
    population = "Population"
  ))) +
  theme_minimal() +
  labs(y = "Total 10-Year Budget (in 000s)")

```

[@fig-growth-related-budget] presents the relationship between average household income, population, and the total 10-year budget allocation for Growth-Related projects. The linear regression line for average household income remains relatively flat, hovering around \$100,000. This suggests that budget allocations for Growth-Related projects are not significantly influenced by the average household income. In other words, funds for these projects are distributed more uniformly across wards, regardless of their income levels. A similar pattern can be observed in [@fig-state-good-repair-budget] where the budget allocation is not strongly correlated with average household income.

In contrast, the linear regression line for population tells a different story. It shows a positive correlation between population size and budget allocation for Growth-Related projects. Wards with higher population densities, such as Ward 10 (Spadina-Fort York), Ward 13 (Toronto Centre), Ward 5 (York South-Weston), and Ward 2 (Etobicoke North), receive larger portions of the Growth-Related budget. This indicates that population density plays a key role in determining how much funding a ward receives for Growth-Related projects, with more populous wards receiving a greater share of the budget.

## Relationship between Average Household Income & Population with Budget Allocation for Service Improvement and Enhancement

```{r}
#| echo: false
#| message: false
#| warning: false
#| fig.cap: "Service Improvement and Enhancement by Ward"
#| label: fig-service-improvement-budget
#| fig-pos: "h"
#| fig.width: 10
#| fig.height: 5

# Plot for Average Household Income by Ward
combined_data |>
  filter(category == "Service Improvement and Enhancement") |>
  select(ward_id, average_household_income, population, total_10_year) |>
  pivot_longer(cols = c("average_household_income", "population"),
               names_to = "variable",
               values_to = "value") |>
  ggplot(aes(x = value, y = total_10_year, label = ward_id)) +
  geom_point() +
  geom_text_repel(hjust = -0.5, vjust = 0.5) +
  geom_smooth(method = "lm", color = "blue", se = TRUE) +
  facet_wrap(~variable, scales = "free", labeller = as_labeller(c(
    average_household_income = "Average Household Income",
    population = "Population"
  ))) +
  theme_minimal() +
  labs(y = "Total 10-Year Budget (in 000s)")

```

[@fig-service-improvement-budget] illustrates the relationship between average household income, population, and the total 10-year budget allocation for Service Improvement and Enhancement projects. These projects  aim to enhance the quality of services provided to residents and improve the overall infrastructure of the city [@toronto_improvements].

The relatively flat linear regression lines for both average household income and population suggest that these factors do not strongly influence the allocation of budgets for Service Improvement and Enhancement projects. This reflects the city's effort to ensure equitable distribution of services and infrastructure improvements across all wards, irrespective of their income levels or population densities.

However, some notable outliers are evident in the data. Ward 11 (University-Rosedale), for example, received the highest budget allocation for Service Improvement and Enhancement projects while also ranking as the second-highest in average household income. This may indicate a more targeted approach to allocating funds in higher-income areas, though the specific reasons could be tied to pressing needs such as ongoing infrastructure upgrades or community development initiatives.

Other wards like Ward 3 (Etobicoke-Lakeshore) and Ward 10 (Spadina-Fort York), which have high population densities, also received significant budget allocations, suggesting that these areas are priorities for service improvements and infrastructure enhancement. 

Despite these outliers, the majority of wards received similar budget allocations for Service Improvement and Enhancement projects, underscoring a city-wide strategy to enhance services uniformly for all residents.

# Discussion {#sec-discussion}

In this study, we anticipated that population density would influence budget allocations across categories. However, our findings show that this relationship varies by category. For Growth-Related projects, population density significantly affects budget distribution, with more populous wards receiving a larger share. This indicates that the city is prioritizing infrastructure development in densely populated areas. However, for categories like Health and Safety, population density is not as strong a determinant, and in some cases, wards with high population density receive less funding than expected.

We also found that wealthier wards receive more funding for Health and Safety projects. This suggests that budget allocations in these categories may be influenced by higher property values, which could pressure the city to maintain and protect these areas more intensively. Additionally, wealthier communities may have more political influence, potentially shaping the city's spending priorities.

Budget allocations for Service Improvement and Enhancement projects were relatively consistent across wards, though some exceptions indicate targeted investments in areas with specific ongoing projects or unique needs. These deviations could be driven by immediate infrastructure requirements or political priorities.

One limitation of this study is the lack of detailed data on specific projects within each budget category, which could provide further insights into why certain wards receive more or less funding. Additionally, while this analysis focused on population density and average household income, other factors like infrastructure needs, community priorities, and political considerations likely play a role in budget decisions. Future research should explore these variables to build a more complete picture of how municipal budgets are allocated. In terms of policy recommendations, the city should adopt equity-based budgeting to ensure underfunded, high-density wards receive adequate resources, particularly for essential services like health and safety. Toronto could also implement a baseline funding model for high-need areas and enhance public engagement in the budget process, ensuring that residents of under served wards have a voice in resource allocation. These actions would promote more transparent, balanced, and equitable urban governance.

In conclusion, this study highlights the complexity of budget allocation in Toronto’s wards. Population density and income levels both play roles in shaping budget distributions, though not always in expected ways. Understanding these dynamics can help policymakers make more equitable resource allocation decisions that better address the needs of all residents.

\newpage

# References
